Explore the SingleR results

### 2025-07-16 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/singler-results.Rmd
all the SingleR into backticks, and programmable table

This notebooks explores the results from running cell type annotation with `SingleR` using the NBAtlas.
The NBAtlas reference was aggregated with `SingelR` prior to model training.

@@

# merged SCE file
sce_file <- file.path(
  merged_dir,
  "SCPCP000004_merged.rds"
)

@@

# Recode NBAtlas cell types where possible so they match with colors
celltype_recoded_df <- celltype_df |>
  # rename to make annotation sources more clear
  dplyr::rename(
    "consensus_validation" = validation_group_annotation,
    "singler" = pruned.labels
  ) |>
  # pivot longer for wrangling
  tidyr::pivot_longer(
    c(consensus_validation, singler),
    names_to = "annotation_type",
    values_to = "label"
  ) |> 
  harmonize_celltypes(label, label_recoded) |>
  dplyr::select(-label)
```

@@

The table below shows the `NBAtlas` cell types and how they are represented in visualizations, unless otherwise stated.

| Broad label shown in figures | Associated NBAtlas labels                                       |
| ---------------------------- | --------------------------------------------------------------- |
| `T cell`                     | `CD4+ T cell`, `CD8+ T cell`, `Treg`, `NKT cell`                |
| `dendritic cell`             | `cDC1`, `cDC2/DC3`, `Migratory cDC`                             |
| `natural killer cell cell`   | `Circulating NK cell`, `TOX2+/KIT+ NK cell`, `Resident NK cell` |
| `monocyte`                   | `CD4+ T cell`, `Classical monocyte`, `Patrolling monocyte`      |
| `myeloid`                    | `Neutrophil`                                                    |
| `Neuroendocrine`             | `Neuroendocrine`                                                |
| `fibroblast`                 | `Fibroblast`                                                    |
| `macrophage`                 | `Macrophage`                                                    |
| `B cell`                     | `B cell`                                                        |
| `plasma cell`                | `Plasma`                                                        |
| `endothelial`                | `Endothelial`                                                   |
| `Stromal other (NBAtlas)`    | `Stromal other`                                                 |
| `Schwann`                    | `Schwann`                                                       |
| `RBCs`                       | `RBCs`                                                          |
| `Immune cycling`             | `Immune cycling`                                                |
| `pDC`                        | `pDC`                                                           |

Cells labeled `singler_unknown` are those for which `SingleR` could not reliably assign an annotation.

@@

This section compares SingleR annotations to consensus cell type annotations using a heatmap.
The heatmap is colored by Jaccard similarity index.

@@

We broadly see high compatibility between SingleR and consensus labels.
In addition, SingleR mostly classifies the unknown consensus cells as Neuroendocrine, which is consistent with our expectation that these are mostly tumor cells.

@@

This section visualizes and compares SingleR annotations to consensus cell type annotations using UMAPs.
Note that the displayed UMAP is from a merged object that has _not been batch-corrected._

Cell type labels have been harmonized between sources wherever possible.
Note that each set of labels has its own "stromal" category which the labels distinguish.
In addition, cells labeled `unknown_consensus` are those with no assigned consensus label, and cells labeled `unknown_singler` are those where SingleR could not confidently assign a label.

@@

First, we display the consensus and SingleR annotations for all cells.

@@

Below we display a faceted UMAP to highlight the SingleR annotations.
Light gray cells in each panel represent all other cell types.

@@

Below, we display faceted UMAPs highlighting a single cell type but considering only cell types that have direct correspondence between SingleR and consensus cell types.
This allows us to see if the normal cells that SingleR is labeling correspond well to the normal cells identified by consensus labels.
Each row displays a cell type where the left column shows the consensus version and the right column shows the SingleR version.

In addition, we include a category below `putative-tumor` to directly compare the unknown consensus labels with the `Neuroendocrine` cells labeled by SingleR. 
While these categories are not necessarily directly comparable, they are each most likely to contain tumor cells.

Also, note that the `myeloid`-labeled SingleR cells are _only_ neutrophils, but the consensus myeloid grouping contains additional cell types.

@@

There appears to be a reasonable correspondence between consensus- and SingleR-colored UMAPS for each cell type.
We see that SingleR classified more cells compared to consensus, which we expected, and we also see that the additional cells that SingleR labeled are roughly in the same neighborhood as their corresponding consensus labels.

@@

* Marker genes identified in NBAtlas corresponding to the atlas cell types
  * This will tell us if we are picking up comparable signal in our data that plays well with NBAtlas
* Consensus cell type marker genes
  * This will provide an independent assessment of the reliability of SingleR normal cell type assignments

@@

This plot does not show the recoded SingleR labels but rather all labels that have associated marker genes; therefore, there are more cell type categories in this plot than in other plots in this report.

@@

This dotplot shows expression of consensus validation marker genes across matching cell types labeled with SingleR.

@@

* `matching annotations`: Cells whose consensus annotation matches the SingleR annotation
* `different annotations`: Cells whose consensus annotation differs from the SingleR annotation
* `consensus not classified`: Cells where the consensus annotation is unknown

Each panel contains only cells that SingleR annotated with the given label, and shows only marker genes associated with that given cell type. 
We expect that cells whose consensus and SingleR annotations agree may show higher gene expression for the given marker.

@@

In most cases, gene expression for cells with matching consensus/SingleR annotations is higher or about the same as is gene expression for cells with different annotations.

>>>>

This notebooks explores the results from running cell type annotation with `SingleR` using the NBAtlas.
The NBAtlas reference was aggregated with `SingleR` prior to model training.

@@

# merged SCE file
sce_file <- file.path(
  data_dir,
  "SCPCP000004_merged.rds"
)

@@

# Recode NBAtlas cell types where possible so they match with colors
celltype_recoded_df <- celltype_df |>
  # rename to make annotation sources more clear
  dplyr::rename(
    "consensus_validation" = validation_group_annotation,
    "singler" = pruned.labels
  ) |>
  # pivot longer for wrangling
  tidyr::pivot_longer(
    c(consensus_validation, singler),
    names_to = "annotation_type",
    values_to = "label"
  ) |> 
  harmonize_celltypes(label, label_recoded)
```

@@

The table below shows the `NBAtlas` cell types and how they are represented in visualizations, unless otherwise stated.
Cells labeled `singler_unknown` are those for which `SingleR` could not reliably assign an annotation.

```{r}
celltype_recoded_df |>
  dplyr::filter(annotation_type == "singler") |>
  dplyr::select(label, label_recoded) |>
  dplyr::distinct() |>
  dplyr::group_by(label_recoded) |>
  dplyr::summarize(all_labels = paste(label, collapse = ", ")) |>
  dplyr::arrange(label_recoded) |>
  dplyr::rename(
    "Label shown in figures" = label_recoded,
    "Underlying NBAltas label(s)" = all_labels
  ) |>
  knitr::kable()
```

@@

This section compares `SingleR` annotations to consensus cell type annotations using a heatmap.
The heatmap is colored by Jaccard similarity index.

@@

We broadly see high compatibility between `SingleR` and consensus labels.
In addition, SingleR mostly classifies the unknown consensus cells as Neuroendocrine, which is consistent with our expectation that these are mostly tumor cells.

@@

This section visualizes and compares `SingleR` annotations to consensus cell type annotations using UMAPs.
Note that the displayed UMAP is from a merged object that has _not been batch-corrected._

Cell type labels have been harmonized between sources wherever possible.
Note that each set of labels has its own "stromal" category which the labels distinguish.
In addition, cells labeled `unknown_consensus` are those with no assigned consensus label, and cells labeled `unknown_singler` are those where `SingleR` could not confidently assign a label.

@@

First, we display the consensus and `SingleR` annotations for all cells.

@@

Below we display a faceted UMAP to highlight the `SingleR` annotations.
Light gray cells in each panel represent all other cell types.

@@

Below, we display faceted UMAPs highlighting a single cell type but considering only cell types that have direct correspondence between `SingleR` and consensus cell types.
This allows us to see if the normal cells that `SingleR` is labeling correspond well to the normal cells identified by consensus labels.
Each row displays a cell type where the left column shows the consensus version and the right column shows the `SingleR` version.

In addition, we include a category below `putative-tumor` to directly compare the unknown consensus labels with the `Neuroendocrine` cells labeled by `SingleR.` 
While these categories are not necessarily directly comparable, they are each most likely to contain tumor cells.

Also, note that the `myeloid`-labeled `SingleR` cells are _only_ neutrophils, but the consensus myeloid grouping contains additional cell types.

@@

There appears to be a reasonable correspondence between consensus- and `SingleR`-colored UMAPS for each cell type.
We see that `SingleR` classified more cells compared to consensus, which we expected, and we also see that the additional cells that `SingleR` labeled are roughly in the same neighborhood as their corresponding consensus labels.

@@

* Marker genes identified in NBAtlas corresponding to the atlas cell types
  * This will tell us if we are picking up comparable signal in our data that plays well with NBAtlas
* Consensus cell type marker genes
  * This will provide an independent assessment of the reliability of `SingleR` normal cell type assignments

@@

This plot does not show the recoded `SingleR` labels but rather all labels that have associated marker genes; therefore, there are more cell type categories in this plot than in other plots in this report.

@@

This dotplot shows expression of consensus validation marker genes across matching cell types labeled with `SingleR`.

@@

* `matching annotations`: Cells whose consensus annotation matches the `SingleR` annotation
* `different annotations`: Cells whose consensus annotation differs from the `SingleR` annotation
* `consensus not classified`: Cells where the consensus annotation is unknown

Each panel contains only cells that `SingleR` annotated with the given label, and shows only marker genes associated with that given cell type. 
We expect that cells whose consensus and `SingleR` annotations agree may show higher gene expression for the given marker.

@@

In most cases, gene expression for cells with matching consensus/`SingleR` annotations is higher or about the same as is gene expression for cells with different annotations.
